Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-05-30 survey/survey.tex
working on sec 3

Instead of spaces consisting solely of unstructured isolated points,
we now investigate structured spaces built from sums and products of
elementary spaces. This structure corresponds to the building blocks
of type theory which are: the empty type ($\bot$), the unit type
($\top$), the sum type ($\uplus$), and the product ($\times$)
type. Before getting into the formal theory, let's consider possible
deformations on the space
$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
product of two subspaces: the subspace $(\top \uplus \bot)$ which
itself is the sum of the space $\top$ containing one element
$\texttt{tt}$ and the empty space $\bot$ and the subspace
$(\top \uplus \top)$ which is the sum of two spaces each containing
the one element $\texttt{tt}$. First, as discussed in the previous
section, any deformation of this space must at least preserve the
number of elements: we can neither create nor destroy points during
any continuous deformation. Seeing that the number of elements in our
example space is 2, a reasonable hypothesis is that we can deform the
space above to any other space with 2 elements such as
$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line, change the orientation of the grid by exchanging
the axes, as well as do other transformations to be discussed in this
section that preserve the number of points.

@@

\begin{defn}
\label{def:rig}
A \emph{commutative semiring} (sometimes called a \emph{commutative
  rig} (commutative ri\emph{n}g without negative elements)
$(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
$R$ named 0 and 1, and two binary operations~$+$ and $\cdot$,
satisfying the following relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{defn}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set''.  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except In homotopy type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

@@

Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second.

>>>>

Instead of spaces consisting solely of unstructured isolated points,
we now investigate structured spaces built from sums and products of
elementary spaces. This structure corresponds to the building blocks
of type theory which are: the empty type ($\bot$), the unit type
($\top$), the sum type ($\uplus$), and the product ($\times$)
type. Before getting into the formal theory, let's consider possible
deformations on the space
$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
product of two subspaces: the subspace $(\top \uplus \bot)$ which
itself is the sum of the space~$\top$ containing one element
$\texttt{tt}$ and the empty space $\bot$ and the subspace
$(\top \uplus \top)$ which is the sum of two spaces each containing
the one element $\texttt{tt}$. First, as discussed in the previous
section, any deformation of this space must at least preserve the
number of elements: we can neither create nor destroy points during
any continuous deformation. Seeing that the number of elements in our
example space is 2, a reasonable hypothesis is that we can deform the
space above to any other space with 2 elements such as
$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line, change the orientation of the grid by exchanging
the axes, as well as do other transformations to be discussed in this
section that preserve the number of points.

@@

\begin{defn}
\label{def:rig}
A \emph{commutative semiring} (sometimes called a \emph{commutative
  rig} (commutative ri\emph{n}g without negative elements))
$(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
$R$ named 0 and 1, and two binary operations~$+$ and $\cdot$,
satisfying the following relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{defn}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

@@

Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second. The
language of permutations lacked the ability to combine permutations by
taking sums and products which led to the awkward non-compositional
programming style illustrated in the full adder example. In this
language the full adder can be programmed as follows:

\amr{***HERE***}
